---
title: "R Notebook"
output: html_notebook
---


Preprocess the raw data to easily import into Matlab for further processing. The final files will be CSV and will be across all specialties, and across Scotland.

The following will process data using the default categorisation (i.e. elective type either inpatient/day case or outpatients)
```{r}
rm(list=ls())
library(dplyr)
library(lubridate)
waiting <- read.csv('data/OngoingWaits_May2024.csv')
capacity <-read.csv('data/CompletedWaits_May2024.csv')
additions <-read.csv('data/AdditionsRemovals_May2024.csv')

waiting_selected <-waiting%>%
  mutate(date=ymd(MonthEnding))%>%
  filter(HBT=="S92000003")%>%
  filter(Specialty=='Z9')%>%
  arrange(date)

capacity_selected <-capacity%>%
  mutate(date=ymd(QuarterEnding))%>%
  filter(HBT=="S92000003")%>%
  filter(Specialty=='Z9')%>%
  arrange(date)

additions_selected <-additions%>%
  mutate(date=ymd(QuarterEnding))%>%
  filter(HBT=="S92000003")%>%
  filter(Specialty=='Z9')%>%
  arrange(date)


write.csv(capacity_selected,file='capacity_electiveType.csv')
write.csv(waiting_selected,file='waiting_electiveType.csv')
write.csv(additions_selected,file='additions_electiveType.csv')

```

Now process the distribution data (again while keeping the inpatient/day case and outpatient separate)
```{r}
rm(list=ls())
library(dplyr)
library(lubridate)
waiting_distribution <-read.csv('data/Distribution_OngoingWaits_May2024.csv')


# distribution for across all specialties, and across Scotland
waiting_distribution_electiveType <-waiting_distribution%>%
  mutate(date=ymd(MonthEnd))%>%
  filter(HBT=="S92000003")%>%
  filter(Specialty=='Z9')%>%
  arrange(date)

write.csv(waiting_distribution_electiveType,file='waiting_distribution_elective.csv')



#### Let us get filtered data for stratified analyses for waiting

# By Region
waiting_distribution_selected_region <-waiting_distribution%>%
  mutate(date=ymd(MonthEnd))%>%
  filter(Specialty=='Z9')%>%
  arrange(date)

# By Specialty
waiting_distribution_selected_specialty <-waiting_distribution%>%
  mutate(date=ymd(MonthEnd))%>%
  filter(HBT=="S92000003")%>%
  arrange(date)

write.csv(waiting_distribution_selected_region,file='waiting_distribution_selected_region.csv')
write.csv(waiting_distribution_selected_specialty,file='waiting_distribution_selected_specialty.csv')


## now process and save the capacity distribution file
capacity_distribution <-read.csv('data/Distribution_CompletedWaits_May2024.csv')


# distribution for across all specialties, and across Scotland
capacity_distribution_electiveType <-capacity_distribution%>%
  mutate(date=ymd(QuarterEnding))%>%
  filter(HBT=="S92000003")%>%
  filter(Specialty=='Z9')%>%
  arrange(date)

write.csv(capacity_distribution_electiveType,file='capacity_distribution_electiveType.csv')



```

Now, import the overall ongoing waiting, and the additions/removals when combining the elective types to get an overall picture
```{r}
rm(list=ls())
library(dplyr)
library(lubridate)
waiting <- read.csv('data/OngoingWaits_May2024.csv')
capacity <-read.csv('data/CompletedWaits_May2024.csv')
additions <-read.csv('data/AdditionsRemovals_May2024.csv')

waiting_overall <-waiting%>%
  mutate(date=ymd(MonthEnding))%>%
  filter(HBT=="S92000003")%>%
  filter(Specialty=='Z9')%>%
  group_by(date)%>%
  mutate(NumberWaiting=sum(NumberWaiting))%>%
  mutate(NumberWaitingOver12Weeks=sum(NumberWaitingOver12Weeks))%>%
  filter(row_number()==1)%>%
  select(-PatientType)%>%
  arrange(date)

capacity_overall <-capacity%>%
  mutate(date=ymd(QuarterEnding))%>%
  filter(HBT=="S92000003")%>%
  filter(Specialty=='Z9')%>%
  group_by(date)%>%
  mutate(NumberSeen=sum(NumberSeen))%>%
  mutate(WaitedOver12Weeks=sum(WaitedOver12Weeks))%>%
  filter(row_number()==1)%>%
  select(-PatientType)%>%
  arrange(date)

additions_overall <-additions%>%
  mutate(date=ymd(QuarterEnding))%>%
  filter(HBT=="S92000003")%>%
  filter(Specialty=='Z9')%>%
  group_by(date)%>%
  mutate(Additions=sum(Additions))%>%
  mutate(Removals=sum(Removals))%>%
  filter(row_number()==1)%>%
  select(-PatientType)%>%
  arrange(date)


write.csv(waiting_overall,file='waiting_overall.csv')
write.csv(capacity_overall,file='capacity_overall.csv')
write.csv(additions_overall,file='additions_overall.csv')

```

Now process the distribution data but combine the elective types to get an overall picture
```{r}
rm(list=ls())
library(dplyr)
library(lubridate)

################# Waiting 

waiting_distribution <-read.csv('data/Distribution_OngoingWaits_May2024.csv')


# distribution for across all specialties, and across Scotland
waiting_distribution_overall <-waiting_distribution%>%
  mutate(date=ymd(MonthEnd))%>%
  filter(HBT=="S92000003")%>%
  filter(Specialty=='Z9')%>%
  group_by(date)%>%
  mutate(LessThan4WeekWait=sum(LessThan4WeekWait))%>%
  mutate(X4To8WeekWait=sum(X4To8WeekWait))%>%
  mutate(X8To12WeekWait=sum(X8To12WeekWait))%>%
  mutate(X12To16WeekWait=sum(X12To16WeekWait))%>%
  mutate(X16To20WeekWait=sum(X16To20WeekWait))%>%
  mutate(X20To24WeekWait=sum(X20To24WeekWait))%>%
  mutate(X24To28WeekWait=sum(X24To28WeekWait))%>%
  mutate(X28To32WeekWait=sum(X28To32WeekWait))%>%
  mutate(X32To36WeekWait=sum(X32To36WeekWait))%>%
  mutate(X36To40WeekWait=sum(X36To40WeekWait))%>%
  mutate(X40To44WeekWait=sum(X40To44WeekWait))%>%
  mutate(X44To48WeekWait=sum(X44To48WeekWait))%>%
  mutate(X48To52WeekWait=sum(X48To52WeekWait))%>%
  mutate(X52To65WeekWait=sum(X52To65WeekWait))%>%
  mutate(X65To78WeekWait=sum(X65To78WeekWait))%>%
  mutate(X78To91WeekWait=sum(X78To91WeekWait))%>%
  mutate(X91To104WeekWait=sum(X91To104WeekWait))%>%
  mutate(X104To117WeekWait=sum(X104To117WeekWait))%>%
  mutate(X117To130WeekWait=sum(X117To130WeekWait))%>%
  mutate(X130To143WeekWait=sum(X130To143WeekWait))%>%
  mutate(X143To156WeekWait=sum(X143To156WeekWait))%>%
  mutate(Over156WeekWait=sum(Over156WeekWait))%>%
  filter(row_number()==1)%>%
  select(-PatientType)%>%
  arrange(date)

write.csv(waiting_distribution_overall,file='waiting_distribution_overall.csv')


################# Capacity

capacity_distribution <-read.csv('data/Distribution_CompletedWaits_May2024.csv')


# distribution for across all specialties, and across Scotland
capacity_distribution_overall <-capacity_distribution%>%
  mutate(date=ymd(QuarterEnding))%>%
  filter(HBT=="S92000003")%>%
  filter(Specialty=='Z9')%>%
  group_by(date)%>%
  mutate(LessThan4WeekWait=sum(LessThan4WeekWait))%>%
  mutate(X4To8WeekWait=sum(X4To8WeekWait))%>%
  mutate(X8To12WeekWait=sum(X8To12WeekWait))%>%
  mutate(X12To16WeekWait=sum(X12To16WeekWait))%>%
  mutate(X16To20WeekWait=sum(X16To20WeekWait))%>%
  mutate(X20To24WeekWait=sum(X20To24WeekWait))%>%
  mutate(X24To28WeekWait=sum(X24To28WeekWait))%>%
  mutate(X28To32WeekWait=sum(X28To32WeekWait))%>%
  mutate(X32To36WeekWait=sum(X32To36WeekWait))%>%
  mutate(X36To40WeekWait=sum(X36To40WeekWait))%>%
  mutate(X40To44WeekWait=sum(X40To44WeekWait))%>%
  mutate(X44To48WeekWait=sum(X44To48WeekWait))%>%
  mutate(X48To52WeekWait=sum(X48To52WeekWait))%>%
  mutate(X52To65WeekWait=sum(X52To65WeekWait))%>%
  mutate(X65To78WeekWait=sum(X65To78WeekWait))%>%
  mutate(X78To91WeekWait=sum(X78To91WeekWait))%>%
  mutate(X91To104WeekWait=sum(X91To104WeekWait))%>%
  mutate(X104To117WeekWait=sum(X104To117WeekWait))%>%
  mutate(X117To130WeekWait=sum(X117To130WeekWait))%>%
  mutate(X130To143WeekWait=sum(X130To143WeekWait))%>%
  mutate(X143To156WeekWait=sum(X143To156WeekWait))%>%
  mutate(Over156WeekWait=sum(Over156WeekWait))%>%
  filter(row_number()==1)%>%
  select(-PatientType)%>%
  arrange(date)

write.csv(capacity_distribution_overall,file='capacity_distribution_overall.csv')

```
